A Three-Step Approach for Assessing Landscape Connectivity via Simulated Dispersal: African Wild Dog Case Study

### 2022-11-29 Chapter_1/04_Manuscript/Manuscript.tex
Incorporated review feedback

We performed all data preparations, spatial computations, and statistical
analysis in {\tt R}, version 3.6.6 \citep{R.2020}. Some helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013, Eddelbuettel.2018}.

@@

\subsection{Step 3 - Connectivity Maps}
\subsubsection{Heatmap}
To identify dispersal hotspots within the study area, we created a heatmap
indicating the absolute frequency at which different areas were traversed by
simulated dispersal trajectories (e.g. \citealp{Hauenstein.2019, Zeller.2020}).
Specifically, we rasterized all simulated trajectories onto a raster with 1 km x
1 km resolution and tallied resulting layers into a single map. This procedure
ensured that every trajectory was only counted once, even if it traversed the
same raster-cell multiple times, thus reducing potential biases caused by
individuals that were surrounded by unfavorable habitat and ``moved in
circles''. To achieve high performance rasterization, we used the R-package {\tt
terra} \citep{Hijmans.2021b}.

@@

\subsubsection{Validation}
To validate our predictions of connectivity, we utilized additional dispersal
data that was collected on eight dispersing coalitions between the years 2019
and 2022 (totalling to 2'668 GPS locations). We used path selection analysis
(PSF, \citealp{Cushman.2010}) to assess if observed dispersal trajectories
followed areas of high predicted connectivity. Similar to SSF, PSF enables to
detect selection for certain features by comparing observed paths to randomly
generated paths. Here, we paired each observed path with 50 random paths that we
generated by randomly rotating and shifting observed paths by a random angle
\(\alpha \sim U(-\pi, +\pi)\) and a random distance \(d \sim U(0 \) km, \(50\)
km\(\)). Along each path, we then extracted connectivity values from the heatmap
(see above) generated after 68, 125, 250, 500, and 2'000 simulated steps,
respectively. Finally, we ran conditional logistic regression to contrast
observed and random paths. In case of systematic selection for high-connectivity
areas, the regression coefficients from the corresponding conditional logistic
regression model should be positive.

@@

% \subsection{Maps}
Each of the three connectivity maps derived from simulated dispersal
trajectories highlighted a different aspect of landscape connectivity. The
heatmap was most suitable for pinpointing frequently traversed areas and showed
that an exceptionally large number of dispersers moved through the regions of
the Moremi NP and the Chobe NP in northern Botswana. \cite{Hofmann.2021}
previously identified the same area as potential dispersal hotspot using LCPA,
however, following their analysis it was not clear whether this was the
consequence of the central location of the region and connections being enforced
between predefined start and endpoints. Contrary to LCPA, a simulation-based
approach as presented here does not require predefined endpoints, as endpoints
emerge naturally from the simulated dispersal trajectories. This is especially
useful for dispersal studies, where known endpoints are usually an unrealistic
assumption \citep{Elliot.2014, Abrahms.2017, Cozzi.2020}. The fact that the same
region was emphasized using vastly different methods to model connectivity thus
reinforces our notion that the area is of exceptional importance to dispersing
wild dogs. Because simulated individuals are not forced to move towards certain
endpoints, a simulation-based approach not only lends itself to study landscape
connectivity, but also to uncover potential dispersal traps
\citep{VanDerMeer.2014} or areas with a high susceptibility for human wildlife
conflicts \citep{Cushman.2018}.

@@

Our approach enabled us to translate a simple set of small-scale behavioral
rules into large scale patterns of connectivity, something previously deemed
computationally unfeasible, yet critical for linking structural and functional
connectivity \citep{Doerr.2011}. Structural connectivity focuses purely on the
spatial arrangement of suitable habitat in the landscape, whereas functional
connectivity also takes into account a species dispersal ability and behavioral
response to the landscape \citep{Tischendorf.2000}. Functional connectivity is
of greater interest to conservation scientists, yet is difficult to quantify
\citep{Baguette.2013}, which is why structural connectivity often serves as
surrogate \cite{Doerr.2011, Fattebert.2015}. LCPA and CT incorporate functional
aspects of connectivity through the permeability surface, which reflects a
species habitat preferences and thus renders behavioral impacts of the lanscape
on the focal species. Aside from rendering habitat preferences, our model also
integrates peculiarities of the focal species movement behavior, thus adding
further insights on functional connectivity. In addition, we could use
independent dispersal data to prove that our predictions of connectiviy aligned
with observed functional connectivity patterns.

@@

% \subsection{Disadvantages of ISSF Simulations}
Despite the many benefits and great flexibility offered by simulations from
ISSFs, one must also be aware of the associated limitations. For example, while
our approach of simulating dispersal proved usedful to assess landscape
connectivity, it was computationally costly. Simulating 80,000 dispersal
trajectories for 2'000 steps across the KAZA-TFCA required five days of
computation on a regular desktop machine (AMD Ryzen 7 2700X octa-core processor
with 3.6 GHz, 64 GB of RAM). The long simulation time was primarily caused by
the massive extent of the study area considered (ca. 1.8 Mio
km\textsuperscript{2}), the large number of simulated trajectories, and the fact
that we extracted covariates along each step, rather than just at their start or
endpoints. Most connectivity studies focus on smaller study areas (e.g.
\citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017, Zeller.2020})
and will therefore require fewer simulations and achieve faster simulation times
(given the same spatial resolution). We also believe that fewer simulated
trajectories will often suffice, as the relative traversal frequency by
simulated trajectories through randomly placed checkpoints across our study area
converged already after 10,500 runs. The exact number of required simulations to
achieve reliable estimates of connectivity will, of course, vary depending on
the structure of the landscape and the dispersal capabilities of the focal
species \citep{Gustafson.1996}. For species that disperse short distances
through homogeneous environments, few simulations may suffice to gauge
connectivity, whereas for species that disperse over long distances through
heterogeneous habitats, a large number of simulations will be required to
sufficiently explore the spectrum of possible routes. Finally, it may often
suffice to extract covariates at each step's start or endpoints, thus
considerably speeding up simulation times \citep{Signer.2017}.

>>>>

We performed all data preparations, spatial computations, and statistical
analysis in {\tt R}, version 4.2.2 \citep{R.2022}. Some helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013, Eddelbuettel.2018}.

@@

\subsection{Step 3 - Connectivity Maps}
\subsubsection{Heatmap}
To identify dispersal hotspots within the study area, we created a heatmap
indicating the absolute frequency at which different areas were traversed by
simulated dispersal trajectories (e.g. \citealp{Hauenstein.2019, Zeller.2020}).
Specifically, we rasterized all simulated trajectories onto a raster with 1 km x
1 km resolution and tallied resulting layers into a single map. This procedure
ensured that every trajectory was only counted once, even if it traversed the
same raster-cell multiple times, thus reducing potential biases caused by
individuals that were surrounded by unfavorable habitat and ``moved in
circles''. To achieve high performance rasterization, we used the R-package {\tt
terra} \citep{Hijmans.2021b}. For a subset of the study area, we also generated
heatmaps at 250 m x 250 m, yet found little qualitative differences to the
coarser resolution, thus suggesting the choice of 1 km x 1 km to be appropriate.

@@

\subsubsection{Validation}
To validate our predictions of connectivity, we utilized additional dispersal
data that was collected on eight dispersing coalitions between 2019 and 2022
(totalling to 2'668 GPS locations). We used path selection analysis (PSF,
\citealp{Cushman.2010}) to assess if observed dispersal trajectories followed
areas of high predicted connectivity. Similar to SSF, PSF enables to detect
selection for certain features by comparing observed paths to randomly generated
paths. Here, we paired each observed path with 50 random paths that we generated
by randomly rotating and shifting observed paths by a random angle \(\alpha \sim
U(-\pi, +\pi)\) and a random distance \(d \sim U(0 \) km, \(50\) km\(\)). Along
each path, we then extracted connectivity values from the heatmap (see above)
generated after 68, 125, 250, 500, and 2'000 simulated steps, respectively.
Finally, we ran conditional logistic regression to contrast observed and random
paths. In case of systematic selection for high-connectivity areas, the
regression coefficients from the corresponding conditional logistic regression
model should be positive.

@@

% \subsection{Maps}
Each of the three connectivity maps derived from simulated dispersal
trajectories highlighted a different aspect of landscape connectivity. The
heatmap was most suitable for pinpointing frequently traversed areas and showed
that an exceptionally large number of dispersers moved through the regions of
the Moremi NP and the Chobe NP in northern Botswana. \cite{Hofmann.2021}
previously identified the same area as potential dispersal hotspot using LCPA,
however, following their analysis it was not clear whether this was the
consequence of the central location of the region and connections being enforced
between predefined start and endpoints. Contrary to LCPA, a simulation-based
approach as presented here does not require predefined endpoints, as endpoints
emerge naturally from the simulated dispersal trajectories. This is especially
useful for dispersal studies, where known endpoints are usually an unrealistic
assumption \citep{Elliot.2014, Abrahms.2017, Cozzi.2020}. The fact that the same
region was emphasized using vastly different methods to model connectivity thus
reinforces our notion that the area is of exceptional importance to dispersing
wild dogs. Because simulated individuals are not forced to move towards certain
endpoints, a simulation-based approach not only lends itself to study landscape
connectivity, but also to uncover potential dispersal traps
\citep{VanDerMeer.2014} or areas with a high susceptibility for human wildlife
conflicts \citep{Cushman.2018}. Using independent dispersal data we showed that
dispersers indeed followed areas of high predicted connectivity. Importantly,
however, these predictions were based on a scenario of a relatively extended
flood, which may not have accurately represented environmental conditions for
dispersers moving through areas affected by the flood. Accounting for such
differences would have improved the predictive performance of our model.

@@

Our approach enabled us to translate a simple set of small-scale behavioral
rules into large scale patterns of connectivity, something previously deemed
computationally unfeasible, yet critical for linking structural and functional
connectivity \citep{Doerr.2011}. Structural connectivity focuses purely on the
spatial arrangement of suitable habitat in the landscape, whereas functional
connectivity also takes into account a species dispersal ability and behavioral
response to the landscape \citep{Tischendorf.2000}. Functional connectivity is
of greater interest to conservation scientists, yet is difficult to quantify
\citep{Baguette.2013}, which is why structural connectivity often serves as
surrogate \citep{Doerr.2011, Fattebert.2015}. LCPA and CT incorporate functional
aspects of connectivity through the permeability surface, which reflects a
species habitat preferences and thus renders behavioral impacts of the lanscape
on the focal species. Aside from rendering habitat preferences, our model also
integrates peculiarities of the focal species movement behavior, thus adding
further insights on functional connectivity. In addition, we successfully used
independent dispersal data to prove that our predictions of connectiviy aligned
with observed functional connectivity patterns.

@@

% \subsection{Disadvantages of ISSF Simulations}
Despite the many benefits and great flexibility offered by simulations from
ISSFs, one must also be aware of the associated limitations. For example, while
our approach of simulating dispersal proved usedful to assess landscape
connectivity, it was computationally costly. Simulating 80,000 dispersal
trajectories for 2'000 steps across the KAZA-TFCA required five days of
computation on a regular desktop machine (AMD Ryzen 7 2700X octa-core processor
with 3.6 GHz, 64 GB of RAM). The long simulation time was primarily caused by
the massive extent of the study area considered (ca. 1.3 Mio
km\textsuperscript{2}), the large number of simulated trajectories, and the fact
that we extracted covariates along each step, rather than just at their start or
endpoints. Most connectivity studies focus on smaller study areas (e.g.
\citealp{Kanagaraj.2013, Clark.2015, McClure.2016, Abrahms.2017, Zeller.2020})
and will therefore require fewer simulations and achieve faster simulation times
(given the same spatial resolution). We also believe that fewer simulated
trajectories will often suffice, as the relative traversal frequency by
simulated trajectories through randomly placed checkpoints across our study area
converged already after 10,500 runs. The exact number of required simulations to
achieve reliable estimates of connectivity will, of course, vary depending on
the structure of the landscape and the dispersal capabilities of the focal
species \citep{Gustafson.1996}. For species that disperse short distances
through homogeneous environments, few simulations may suffice to gauge
connectivity, whereas for species that disperse over long distances through
heterogeneous habitats, a large number of simulations will be required to
sufficiently explore the spectrum of possible routes. Finally, it may often
suffice to extract covariates at each step's start or endpoints, thus
considerably speeding up simulation times \citep{Signer.2017}.
